Numerical simulation of the Cordilleran ice sheet \\ through the last glacial cycle

### 2014-07-08 kappa.tex
Improve glacial history and add more refs.

% ----------------------------------------------------------------------
\subsection{Palaeo-glaciology}
% ----------------------------------------------------------------------

During the last glacial maximum, the Cordilleran ice sheet covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}). The southern margin is the most
extensively studied. It consists of different piedmont lobes that have formed
at the outlet of the main valleys. The Puget lobe was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape known as the Channeled Scabland.

The western margin of the ice sheet covered part of the continental shelf. An
ice cap covering Vancouver Island was well-connected to main body of ice on the
continent. A significant part of the ice sheet thus drained into the Georgia
Strait and engulfed between Vancouver Island and the Olympic Peninsula before
it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean.
Further to the North, the ice margin reached Queen Charlotte Islands, yet the
archipelago may not have been fully covered by ice
flowing from the land. Submarine troughs likely idicate the erosional action of
former ice streams, yet it is not clear whether they have formed during the
last glacial cycle or earlier.

The northern margin of the ice sheet is in several ways more complex than the
southern margin. Firstly, geomorphological evidence indicate multiple glacier
advances. During the last glacial cycle, at least two glaciations occured, and
in some places, the first glaciation was more extensive than the second.
Secondly, geomorphological evidence indicate that the extent of glaciers in the
North was relatively restricted, despite very cold temperatures at present.

Finally, the eastern margin of the ice sheet is the most unclear. Although it
is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
the precise location of this junction is unknown, and little information exists
on its timing. More importantly, the nature of the interaction between the two
ice sheets is uncertain. It is not known for instance, if the junction lasted
long enough for the ice sheets to significantly alter their respective flow
dynamics. Because the present thesis focuses specifically on the Cordilleran
ice sheet, our largest assumption certainly consists of neglecting buttressing
effects from the Laurentide ice sheet.

Although the last glacial maximum extent of the Cordilleran ice sheet, apart of
its eastern margin, is relatively well known, the history of deglaciation is
heavily debated, while little information exists on earlier dynamics of ice
sheet.

>>>>

% ----------------------------------------------------------------------
\subsection{Glacial history}
% ----------------------------------------------------------------------

The first presence of an ice sheet in the Cordilleran mountains is indicated by
glaciofluvial deposits dated at $2.64^{+0.20}/_{-0.18}$\,Ma
\citep{Hidy.etal.2013}. This suggests that the North American Cordillera has
been glaciated many times, with the first glaciations occurring even before the
onset of the 100-ka glacial cycle (\cref{fig:paleo-timeseries}). However, only
the last few glaciations can be reconstructed from the geologic record, whereas
much of the older evidence has been overridden by subsequent glacial
re-advances and other geomorphological processes.

In the northern sector of the ice sheet, at least four glaciations have been
identified
    \citep{Duk-Rodkin.1999, Ward.etal.2007, Ward.etal.2008,
           Briner.Kaufman.2008, Demuro.etal.2012,
           Stroeven.etal.2010, Stroeven.etal.2014}.
The two older ones, often collectively referred to as the Reid glaciations,
pre-date the last glacial cycle, and thus are beyond the scope of this study.
The two younger glaciations, though, occurred both during the last glacial
cycle. First, the Gladstone glaciation reached its maximum between
$50.4\pm3.0$ and $54.3\pm2.0$\,\unit{^{10}Be\,ka} \citep{Ward.etal.2007},
during a period of high global ice volume on Earth, known as the Marine Isotope
Stage (MIS)~4. However, evidence from the Gladstone glaciation has not been
found in the southern part of the ice-covered region, making it difficult to
seize its magnitude. Second, the McConnell glaciation is associated with end
moraines of oldest minimum apparent exposure ages of
$15.7\pm1.5$ and $17.7\pm1.6$\,\unit{^{10}Be\,ka} \citep{Stroeven.etal.2014},
which corresponds to the last global ice volume maximum (MIS~2). In contrary to
the Gladstone glaciation, the McConnell glaciation has been correlated with a
contemporaneous advance of the southern ice sheet margin, known as the Fraser
glaciation, with a maximum extent framed between
$17.4$ and $16.4$\,\unit{^{14}C\,cal\,ka} \citep[Fig.~4]{Porter.Swanson.1998}.
Within errors inherent to the dating methods \citep{Heyman.etal.2011}, the
McConnell and Fraser glaciations correspond to the same glacial episode, which
resulted in the Last Glacial Maximum (LGM) extent of the Cordilleran ice sheet.
This is the period where most geological observations point at.

It is generally thought that the Cordilleran ice sheet was short-lived
    \citep[e.g.]{Clague.etal.1980, Clague.1985, Cosma.etal.2008}.
The ice sheet originated from the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
During the last glacial maximum, ice covered most of the
Canadian Cordillera, and some of the adjacent land in north-western contiguous
USA and Alaska (\cref{fig:map-cordillera}).

The southern margin of the ice sheet has been studied extensively, and is in
several respects better understood than its northern counterpart
\citep{Booth.etal.2003}. It consists of different piedmont
lobes that have formed at the outlet of the main valleys. The Puget lobe
\citep[e.g.]{Thorson.1980, Porter.Swanson.1998} was the westernmost and
largest lobe. It covered the Puget Lowland, a region that currently includes
the cities of Vancouver and Seattle. East of the Cascade Range, were found the
Okanagan and Purcell lobes. The latter once dammed the Clark Fork River to form
Glacial Lake Missoula \citep{Pardee.1910}. As the ice sheet retreated, the lake
repeatedly and catastrophically drained to the Pacific Ocean
\citep{Bretz.1923,Waitt.1980}. These floods
contributed to shape a distinctive landscape, known as the Channeled Scabland.

The western margin of the ice sheet covered part of the continental shelf. An
ice cap covering Vancouver Island was well-connected to main body of ice on the
continent \mref. A significant part of the ice sheet thus drained into the Georgia
Strait and engulfed between Vancouver Island and the Olympic Peninsula where
it formed the Juan the Fuca marine ice lobe, calving into the Pacific Ocean \mref.
Further to the North, the ice margin reached Queen Charlotte Islands, yet the
archipelago may not have been fully covered by ice
flowing from the land. Submarine troughs likely indicate the erosional action of
former ice streams, yet it is not clear whether they have formed during the
last glacial cycle or earlier \mref.

Finally, the eastern margin of the ice sheet is the most unclear. Although it
is known that the Cordilleran ice sheet collided with its Laurentide neighbour,
the nature of the interaction between the two ice sheets and the precise
location of this junction is unclear, while little information exists on its
timing
    \citep[e.g.][]{Jackson.etal.1997, Bednarski.Smith.2007}.
Because the
present thesis focuses specifically on the Cordilleran ice sheet, our largest
assumption certainly consists of neglecting buttressing effects from the
Laurentide ice sheet.

The ice sheet is believed to have retreated quickly. Most of the evidence comes
from the Fraser Plateau, where it indicates that some of the highest relief
emerged from the ice prior to the lowlands, possibly resulting in thin,
stagnant ice remnants in the interior
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
Much less is known about ice sheet decay in the northern parts of the
ice-covered areas, but glacial landforms from the Liard Lowland suggest a
different mode of deglaciation, with active retreat of the ice lobe towards the
surrounding mountain ranges \citep{Margold.etal.2013}. Patterns of
deglaciation, as well as the possibility for a late-glacial re-advance, are
discussed in further detail and compared with modelling results in \CCYC.
